How to select predictive models for decision-making or causal inference?

### 2024-09-01 how_to_select_latex.tex
iter R2

Given complex health data, which predictive model is to be most trusted to yield
valid causal estimates needed to motivate individual treatment decisions?
As no single machine-learning method performs best across all data sets, there is a
pressing need for clear guidelines to select outcome models for causal
inference.

@@

Similarly to the ATE, at the individual level, the CATE:
\begin{equation}
    \tau(x) = \mu_{1}(x) - \mu_{0}(x)
    \label{eq:cate_estimate}
\end{equation}

\paragraph{Robinson decomposition}
G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}. It introduces two new statistical estimates, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.

@@

\subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, g-computation itself does not control error on both populations as highlighted in \ref{fig:toy_example}. Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}. Thus, selection procedures, more elaborated than MSE are needed to favor such models. A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ on the observed data $O$ and a splitting strategy of the data.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

@@

\subsection{Estimation and model selection procedure}\label{problem:estimation_procedure}

@@

The simulation has been designed to control precisely the overlap parameter. The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions than simulated covariates. These simulations cover a wide range of causal settings, such ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.

@@

Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
nuisance estimates is a form of double dipping which can lead errors in
nuisances correlated to that of outcome models
\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
selection and, strictly speaking, push
to split out a third separated data set --a ``nuisance set''-- to fit the
nuisance models. The drawback is that it depletes the data available for
model estimation and selection. However, Figure
\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
nuisance set and the simpler shared nuisance-candidate set procedure.

>>>>

Given complex health data, which predictive model is to be most trusted to yield
valid causal estimates needed to motivate individual treatment decisions?
As no single machine-learning method performs best across all datasets, there is a
pressing need for clear guidelines to select outcome models for causal
inference.

@@

Similarly to the ATE, at the individual level Eq.\ref{eq:mu_identification} links the CATE to statistical quantities:
\begin{equation}
    \tau(x) = \mu_{1}(x) - \mu_{0}(x)
    \label{eq:cate_estimate}
\end{equation}

\paragraph{Robinson decomposition}
G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}. It introduces two new statistical estimates, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

\subsection{Model-selection, oracle and feasible risks}\label{sec:problem:model_selection}

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, g-computation itself does not control error on both populations as highlighted in Figure \ref{fig:toy_example}. Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}. Thus, selection procedures, more elaborated than MSE are needed to favor the best model for each dataset. A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

@@

\subsection{Model selection procedure}\label{problem:estimation_procedure}

@@

The simulations have been designed to evaluate the effect of the overlap parameter. The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions. These simulations cover a wide range of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.

@@

Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
nuisance estimates is a form of double dipping which can lead errors in
nuisances correlated to that of outcome models
\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
selection and, strictly speaking, push
to split out a third separated dataset --a ``nuisance set''-- to fit the
nuisance models. The drawback is that it depletes the data available for
model estimation and selection. However, Figure
\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
nuisance set and the simpler shared nuisance-candidate set procedure.
